---
title: "Data censoring in social sciences"
author: "Bikramjit Das"
date: "2025"
output: 
 html_document: 
    toc: true
    toc_depth: 2
    toc_float: true
    highlight: tango
geometry: margin=1in
fontsize: 12pt
linkcolor: blue    
---

# Data preparation

First, we prepare the working environment, load the survival package and `extramarital.csv` data.
```{r}
rm(list=ls()) # Remove all variables from the R environment to create a fresh start
# setwd("...")  # Setup the working directory
```

### Installation

Next we install the `survival` package, which includes functions for running the Tobit model.

```{r}
if(!require(survival)){
  install.packages("survival")}
library(survival)
```  

Next we read the "extramarital.csv" file.
```{r}
exm <- read.csv("extramarital.csv")
str(exm)
```

The dataset consists of 6366 observations of 7 variables:

 - `marriage_rating`: 5 = very good, 1 = very poor
 - `age`: 17.5 = under 20, 22 = 20 to 24, ..., 42 = 40 or over 
 - `yrs_married`: 0.5 = less than 1 year, 2.5 = 1 to 4 years ...
 - `religiosity`: 4 = strong, ..., 1 = no
 - `education`: level of education, 9 = grade school, 12 = high school, ... 
 - `occupation`: 6 = professional with advanced degree, 5 = managerial, administration, ... 1 = student
 - `time_in_affairs`: amount of time spent in extra-marital affairs

The dataset was collected in a survey by Redbook and analyzed by Ray C. Fair (Yale).

# Modeling and Analysis

##  Modelling 
  Let's take a better look at the data
```{r}  
table(exm$time_in_affairs>0)
```
We observe that about 32% of people in the dataset are having an affair. 
<!-- # FALSE  TRUE  -->
<!-- # 4313  2053  -->

 Let's look at the distribution of `time_in_affairs`.
```{r}
hist(exm$time_in_affairs,100)
```

 Note that in this dataset there are a large number of zeros for the affair variable. If someone is not having an affair, a small decrease in one of the predictors may not change his/her  chances of having an affair in comparison to a person having an affair. Hence, one needs to account for possibly two types of people: those who are unlikely to be moved away from 0 by modest changes in the predictors, and those who are not 0 and likely to change by modest change in predictors.
To do this, we use the *Tobit* model.

## Training and test dataset

We create a training and a test dataset first.
```{r}
set.seed(100)
spl <- sample(nrow(exm),0.7*nrow(exm))
train <- exm[spl,]
test <- exm[-spl,]
str(train)
str(test)
```

## Tobit model
Now we fit the  Tobit model.
```{r}
?Surv     # Create a survival object, usually used as a response variable in a model formula.
?survreg  # Fit a parametric survival regression model.
model1 <- survreg(Surv(time_in_affairs,time_in_affairs>0,type="left")~.,data=train,dist="gaussian")
```

 This fits a Tobit model where the data are left censored at zero. The error terms are assumed to be Gaussian (normally distributed). Let us look at the model summary.
```{r}
 summary(model1)
```

Now, let's look at the estimated parameters ($\beta$-coefficients. )
 
 - `marriage_rating`: negative, meaning that as the marriage rating increases, the chances/length of an affair decrease.
 - `occupation`: positive, meaning that there are greater chances of an affair with a better occupation.
 - `yrs_married`: positive
 - `religiosity`: negative

 Also, note that all coefficients are fairly significant. The model has a log likelihood of -5399.9.

### Prediction
To check how good our model fits, we predict on the test data 

```{r}
predict1 <- predict(model1,newdata=test) # Prediction provides the latent predicted values of the variables (positive and negative)
# plot(predict1)
```

## Benchmark model

We also us a benchmark model, namely *linear regression* to compare.
```{r}
model2 <- lm(time_in_affairs~.,data=train)
summary(model2)
predict2 <- predict(model2,newdata=test)
# plot(predict2)
```


## Model comparison

Now let's compare the two models
```{r}
table(predict1 <= 0, test$time_in_affairs==0)
```
<!-- #       FALSE TRUE -->
<!-- # FALSE    129   48 -->
<!-- # TRUE    473 1260 -->
```{r}
table(predict2 <= 0, test$time_in_affairs==0)
```
<!-- #       FALSE  TRUE -->
<!-- # FALSE   571 1184 -->
<!-- # TRUE    31   124 -->
<!-- # -->

Clearly, the Tobit model has better accuracy in terms of accounting for  the large number of individuals for whom the extramarital affairs variable was 0.

Note that here we are simply comparing how many of the $\beta_0+\sum_j \beta_j x_{ij}$ values are below and above zero. One can also be more careful and compute
$$\mathbb{P}(Y_i>0)=\mathbb{P}(Y_i^*>0), \quad\quad \text{and, }  \quad\quad \mathbb{P}(Y_i=0)=\mathbb{P}(Y_i^*\le0).$$


##### End of Data censoring for Social sciences analyses.

