---
title: "Recommendation systems 1"
subtitle: "Clustering techniques"
author: "Bikramjit Das"
date: "2025"
output: 
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    toc: yes
    toc_depth: 2
    number_sections: yes
---

First, we prepare the working environment.
```{r}
rm(list=ls()) # Clear the environment
# setwd("...")  # Setup the working directory
```


# MovieLens dataset

The data come from [grouplens](https://grouplens.org), a research lab in the Department of Computer Science and Engineering at the University of Minnesota.

We begin by loading the data. Because the dataset contains strings, we use the option `stringsAsFactors=FALSE` (so as to make sure that character vectors are NOT converted into factors). The dataframe consists of two variables with 8569 movies with movieID and title.

```{r}
movies <- read.csv("movies.csv",stringsAsFactors=FALSE)
str(movies)
```

A quick look at the dataframe:

```{r}
head(movies) # First part of the dataframe
summary(movies) # Summary of the data
```

The movie genre is contained in a different file. To read it, we use the following commands:

```{r}
# Count the number of fields separated by "|" in each row of genres.csv
countfields <- count.fields("genres.csv", sep="|") 
# The first movie, for example has 5 genres listed
countfields[1] 
# With this command, we can see that movies have from 1 to 7 genres listed
min(countfields)
max(countfields)
# Now, we load genres.csv. The command creates a dataframe with 8569 observations and 7 variables with column names X1 ... X7.
genres <- read.csv("genres.csv", header=FALSE, sep="|", col.names=c("X1","X2","X3","X4","X5","X6","X7"), stringsAsFactors=TRUE)
# Note that each variable has different number of levels
str(genres) 
```
 
For this class, we need to work with a matrix having 8,569 rows and a number of columns equal to the number of genres. For each cell, we want to know whether a certain genre is associated (1) or not (0) to a certain movie. This requires some processing of the data contained in `genres`. We use a dedicated function to do it.

```{r}
# Source and run the function
source('process_genre_data.R')
M <- process_genre_data(genres)
```

Last step, we create a dataframe (named `Data`) with movie title and classification.

```{r}
Data <- as.data.frame(M)
Data$title <- movies$title
Data <- Data[,-19] # Drops the 19th column, which corresponds to the "" category
```

We are now ready for our clustering exercise.


# Hierarchical clustering

To run hierarchical clustering in R, we must follow two steps. First, we use the function `dist` to calculate the distance between movies using the first 19 columns (genres of movies).

```{r}
distances <- dist(Data[,1:19], method="euclidean")
dim(Data)
# length(distances)
```

Note that we computed `r 8569*8568/2` distances.

Then, we execute hierarchical clustering. We use Ward's distance method to find compact clusters.

```{r}
clusterMovies1 <- hclust(distances, method="ward.D2")
```

Plot the dendrogram. We have several movies, so the lists at the bottom cannot be read.

```{r}
plot(clusterMovies1) 
```

Let's now cut the dendrogram into 10 clusters and compute the average value across the clusters for the Action variable. Higher value indicates many movies in the clusters are action movies.

```{r}
clusterGroups1 <- cutree(clusterMovies1, k=10)
tapply(Data[,1], clusterGroups1, mean) 
```

Create a matrix "Cat1" where rows denote categories and columns indicate clusters

```{r}
Cat1 <- matrix(0,nrow=19,ncol=10)
for(i in 1:19)
{Cat1[i,] <- tapply(Data[,i], clusterGroups1, mean)}
rownames(Cat1) <- colnames(Data)[1:19]
Cat1
```

Looking at `Cat1`, we can infer that the 10 clusters are organized as follows:
  
* Cluster 1: Fantasy, Comedy, Children, Adventure
* Cluster 2: Romance, Comedy, Drama
* Cluster 3: Comedy
* Cluster 4: Action, Drama, Thriller, Crime
* Cluster 5: Thriller, Horror
* Cluster 6: Drama
* Cluster 7: Drama, Romance
* Cluster 8: Drama, Thriller
* Cluster 9: Action, Sci-fi
* Cluster 10: Documentary

Let's take a look at various movies

```{r}
# subset(Data$title, clusterGroups1==6)
subset(Data, movies$title=="Grand Budapest Hotel, The (2014)")
clusterGroups1[8418]
subset(Data, movies$title=="Moneyball (2011)")
clusterGroups1[7925]
subset(Data, movies$title=="John Wick (2014)")
clusterGroups1[8514]
subset(Data, movies$title=="Star Trek (2009)")
clusterGroups1[7293]

```


# K-means clustering

This algorithm is not deterministic, so it is best practice to first set the seed.

```{r}
set.seed(1)
```

The following command performs a K-means clustering using k=10 and 20 random initial configurations (a random set of rows is chosen as initial centers).

```{r}
clusterMovies2 <- kmeans(Data[,1:19],centers=10,nstart=20)
```

Total within cluster sum of squares (we want this number to be small):

```{r}
clusterMovies2$tot.withinss
clusterMovies2$totss

```

Let's try again, but, this time, with just one initial configuration:

```{r}
set.seed(1)
clusterMovies3 <- kmeans(Data[,1:19],centers=10,nstart=1)
clusterMovies3$tot.withinss 
clusterMovies3$totss
```

We can see that picking more starting point is useful!

Let's now vary the value of k and plot the error in the cluster as a function of k:

```{r}
set.seed(1)
fit <- 0
for(k in 1:20)
{clusterMovies4 <- kmeans(Data[,1:19], centers=k, nstart=20)
fit[k] <- clusterMovies4$tot.withinss}
plot(1:20,fit)
```

Let's see the corresponding clusters:

```{r}
Cat2 <- matrix(0,nrow=19,ncol=10)
for(i in 1:19)
{Cat2[i,] <- tapply(Data[,i], clusterMovies2$cluster, mean)}
rownames(Cat2) <- colnames(Data)[1:19]
Cat2
```



Looking at `Cat1`, we can infer that the 10 clusters are organized as follows:
  
* Cluster 1: Crime, Drama, Thriller
* Cluster 2: Adventure, Animation, Children, Comedy
* Cluster 3: Drama
* Cluster 4: Documentary
* Cluster 5: Horror
* Cluster 6: Action, Adventure
* Cluster 7: Comedy, (Drama)
* Cluster 8: Action, Thriller
* Cluster 9: Comedy, Romance
* Cluster 10: Horror, Thriller




```{r}

subset(Data, movies$title=="Grand Budapest Hotel, The (2014)")
clusterMovies2$cluster[8418]
subset(Data, movies$title=="Moneyball (2011)")
clusterMovies2$cluster[7925]
subset(Data, movies$title=="John Wick (2014)")
clusterMovies2$cluster[8514]
subset(Data, movies$title=="Star Trek (2009)")
clusterMovies2$cluster[7293]

```








