---
title: "Test your knowledge of Clustering and Collaborative filtering"
date: "Fall 2025"
output:
 html_document: 
    toc: true
    toc_depth: 2
    toc_float: true
    highlight: tango
geometry: margin=1in
fontsize: 12pt
linkcolor: blue    
---

# Question 1 
## a) 
There are a total of 11580 observations.
```{r}
stocks <- read.csv("StocksCluster.csv")
str(stocks)
```

## b)
54.6114 percent of the observations had positive returns in December 

```{r}
table(stocks$PositiveDec)/nrow(stocks)
```

## c)

Maximum correlation = 0.1916727856. By observing the correlation matrix, we can see it is between the returns of October and November.
```{r}
a <- cor(stocks[,1:11])
sort(cor(stocks[,1:11]))
```

## d)
Maximum mean return is in April and minimum mean return is in September.

```{r}
colMeans(stocks[,1:11])
which(colMeans(stocks[,1:11]) == max(colMeans(stocks[,1:11])))
which(colMeans(stocks[,1:11]) == min(colMeans(stocks[,1:11])))
```

## e)

Accuracy = 0.5711818
```{r}
library(caTools)
set.seed(144)
spl <- sample.split(stocks$PositiveDec, SplitRatio = .7)
stocksTrain <- subset(stocks, spl == TRUE)
stocksTest <- subset(stocks, spl == FALSE)
stocksModel <- glm(PositiveDec ~ ., data = stocksTrain, family = "binomial")
stocksPredict <- predict(stocksModel, newdata = stocksTrain, type = "response")
table(stocksPredict >= 0.5, stocksTrain$PositiveDec)
(990+3640)/(990+787+2689+3640)
```
## f)
Accuracy on test set = 0.5670697 (close to what we see on the training set)

```{r}
stocksPredicttest <- predict(stocksModel, newdata = stocksTest, type = "response")
table(stocksPredicttest >= .5, stocksTest$PositiveDec)
(417+1553)/(417+344+1160+1553)
```

## g)

Baseline model predicts positive return (majority in training set). Accuracy of this on test set = 0.5460564. 
```{r}
table(stocksTrain$PositiveDec)/nrow(stocksTrain)
table(stocksTest$PositiveDec)/nrow(stocksTest)
1897/(1897+1577)
```

## h)

We remove the dependent variable in the clustering phase since needing to know the dependent variable in assigning an observation to a cluster defeats the purpose of the methodology (option iii).  

```{r}
limitedTrain <- stocksTrain
limitedTrain$PositiveDec <- NULL
limitedTest <- stocksTest
limitedTest$PositiveDec <- NULL
```

## i)

Mean of ReturnJan in training set = 0 (normalized) while in the test set it is -0.0004185886.

```{r}
#install.packages("caret")
library(caret)
preproc <- preProcess(limitedTrain)

# for (i in ncol(limitedTrain)) {
#   limitedTrain[, i] <- (limitedTrain[, i] - mean(limitedTrain[, i])) / sd(limitedTrain[, i])
# }
# 
# for (i in ncol(limitedTrain)) {
#   limitedTest[, i] <- (limitedTest[, i] - mean(limitedTrain[, i])) / sd(limitedTrain[, i])
# }

normTrain <- predict(preproc, limitedTrain)
normTest <- predict(preproc, limitedTest)
colMeans(normTrain)
colMeans(normTest)
```
## j) 
The mean is 0 in training set since the distribution in training and test set are not identical and normalization is done on the training set. 

## k)
Cluster 2 has the largest number of observations (4731).

```{r}
set.seed(144)
km <- kmeans(normTrain, centers = 3)
table(km$cluster)
```

## l)
Cluster 2 has the 2029 observations in the test set.
```{r}
#install.packages("flexclust")
library(flexclust)
?as.kcca
km.kcca <- as.kcca(km, normTrain)
clusterTrain <- predict(km.kcca)
clusterTest <- predict(km.kcca, newdata = normTest)
table(clusterTest)
table(clusterTrain)

sum((normTest[1] - km$centers[1,])^2)
sum((normTest[1] - km$centers[2,])^2)
sum((normTest[1] - km$centers[3,])^2)
```

## m)
stocksTrain1 (from cluster 1) has the highest average value of the dependent variable in the training set.

```{r}
stocksTrain1 <- subset(stocksTrain, clusterTrain == 1)
stocksTrain2 <- subset(stocksTrain, clusterTrain == 2)
stocksTrain3 <- subset(stocksTrain, clusterTrain == 3)
stocksTest1 <- subset(stocksTest, clusterTest == 1)
stocksTest2 <- subset(stocksTest, clusterTest == 2)
stocksTest3 <- subset(stocksTest, clusterTest == 3)
mean(stocksTrain1$PositiveDec)
mean(stocksTrain2$PositiveDec)
mean(stocksTrain3$PositiveDec)
```

## n)
ReturnFeb, ReturnMar, ReturnJuly, ReturnSep have positive coefficient in one of the three models (at least) and negative coefficient in one of the three models (at least). 

```{r}
StocksModel1 <- glm(PositiveDec ~ ., data = stocksTrain1, family = "binomial")
StocksModel2 <- glm(PositiveDec ~ ., data = stocksTrain2, family = "binomial")
StocksModel3 <- glm(PositiveDec ~ ., data = stocksTrain3, family = "binomial")
summary(StocksModel1)
summary(StocksModel2)
summary(StocksModel3)
```

## o)

Accuracy is 0.6446125, 0.5250126 and 0.625323 respectively.
```{r}
PredictTest1 <- predict(StocksModel1, newdata = stocksTest1, type = "response")
PredictTest2 <- predict(StocksModel2, newdata = stocksTest2, type = "response")
PredictTest3 <- predict(StocksModel3, newdata = stocksTest3, type = "response")
table1 <- table(PredictTest1 >= 0.5, stocksTest1$PositiveDec )
table1
(43+639)/(43+639+26+350)
table2 <- table(PredictTest2 >= 0.5, stocksTest2$PositiveDec )
table2
(227+812)/(227+812+221+719)
table3 <- table(PredictTest3 >= 0.5, stocksTest3$PositiveDec )
table3
(119+123)/(119+123+76+69)
```

## p)

Overall accuracy = 0.5794473

```{r}
AllPredictions <- c(PredictTest1, PredictTest2, PredictTest3)
AllOutcomes <- c(stocksTest1$PositiveDec, stocksTest2$PositiveDec, stocksTest3$PositiveDec)
table4 <- table(AllPredictions >= 0.5, AllOutcomes)
table4
(439+1574)/(439+1574+323+1138)
```

##### End of Exercise