---
title: "Decision trees (Part 1)"
author: "Bikramjit Das"
date: "2025"
output: 
 html_document: 
    toc: true
    toc_depth: 2
    toc_float: true
    highlight: tango
geometry: margin=1in
fontsize: 12pt
linkcolor: blue    
---

First, we prepare the working environment.
```{r}
rm(list=ls())   # Clear the environment
# setwd("...")  # Setup the working directory
```


# Data Preparation

## Loading data
   Load the data on the US Supreme Court
```{r}
supreme <- read.csv("supreme.csv") # We have 623 observations and 20 variables
#str(supreme)       # Internal structure of the dataframe 
head(supreme)    # First part of the dataframe
# summary(supreme) # Summary of the data
```

## Comments
  A few comments on the data

- ```docket```: case number
- ```term```: year the case was discussed
- ```party_1``` and ```party_2```: parties involved in the case
- ```rehndir```, ```stevdir```, ```ocondir```, ```scaldir```, ```kendir```, ```soutdir```, ```thomdir```, ```gindir```, ```brydir```: direction of the judgement of the 9 judges (i.e., Rehnquist, Stevens, O'Connor, Scalia, Kennedy, Souter, Thomas, Ginsburg, Breyer). 0 means liberal, 1 conservative. 9 indicates that the vote is not available for the case.
- ```petit```: petitioner type
- ```respon```: respondent type
- ```circuit```: circuit of origin
- ```uncost```: binary number indicating if the petitioner argued the practice was unconstitutional
- ```lctdir```: lower court direction (liberal or conservative)
- ```issue```: area of issue
- ```result```: 0 means liberal, 1 conservative

## More insights
  A few more insights on the data

Let's now focus on a specific judge, say Stevens. We begin by removing data with no entry from Stevens.
```{r}
stevens <- subset(supreme[,c("docket","term","stevdir","petit","respon","circuit","unconst","lctdir","issue","result")],supreme$stevdir!=9)
```

Then, we process the output result to affirm or reverse for judge Stevens. This creates a new variable (```$rev```) in the dataframe that takes a value of 1 if Stevens' decision reverses the lower court decision and 0 if the judge affirms the lower court decision. Note that a similar analysis can be done for the other judges or for the overall case result.

```{r}
stevens$rev <- as.integer((stevens$lctdir=="conser" & stevens$stevdir==0) | (stevens$lctdir=="liberal" & stevens$stevdir==1))
```

The table below gives us some stats about judge Stevens' decisions. He voted 341 times to reverse the lower court direction and 278 times to affirm it. So this shows that the data set is reasonably balanced.

```{r}
table(stevens$rev)
```

## Train and test

Finally, we split the data into training and testing datasets. To this purpose, we use the ```caTools``` package.

```{r, message=FALSE}
if(!require(caTools)){                          # install or load the package
  install.packages("caTools")
  library(caTools)
}
set.seed(123)                                   # set seed for random sampling
spl <- sample.split(stevens$rev,SplitRatio=0.7) # We use 70% of the data for training
train <- subset(stevens,spl==TRUE);             # training dataset
test <- subset(stevens,spl==FALSE);             # testing dataset
```


# Warm up: Logistic regression

Before introducing the Classification and Regression Trees (CARTs), let's warm up with a simple exercise: predicting justice Stevens' decisions with Logistic Regression.

```{r}
m1 <- glm(rev~petit+respon+circuit+unconst+issue+lctdir,data=train,family="binomial")
summary(m1)
```

Let's now try to make a prediction on the test data. We cannot use the command ```p1 <- predict(m1,newdata=test,type="response")``` on the present version of the data, because ```issue``` has only one realization of the IR value (Interstate Relations), which is not estimated in the training set. Let's thus modify the test dataset as follows:

```{r}
test <- subset(test,test$issue!="IR")
p1 <- predict(m1,newdata=test,type="response")
```

How does the model perform?

```{r}
table(p1>=0.5,test$rev)
```

So the accuracy is ```r sum(diag(table(p1>=0.5,test$rev)))/sum(table(p1>=0.5,test$rev))```. Let's plot the ROC Curve.

```{r, message=FALSE}
if(!require(ROCR)){
  install.packages("ROCR")
  library(ROCR)
}
pred <- prediction(p1,test$rev)
perf <- performance(pred, x.measure="fpr", measure="tpr")
# performance(pred,measure="auc")
as.numeric(performance(pred,measure="auc")@y.values)
plot(perf)
```


# CART
  Now we start with Classification And Regression Trees (CARTs).
```{r, message=FALSE}
if(!require(rpart)){        # rpart is for CART
  install.packages("rpart")
  library(rpart)
}
```

## Building a CART

The command ```cart1 <- rpart(rev~petit+respon+circuit+lctdir+issue+unconst,data=train)```, will not work, because ```rpart``` builds a regression tree by default. So, remember to specify you are working on a classification problem! 

```{r}
cart1 <- rpart(rev~petit+respon+circuit+lctdir+issue+unconst,data=train,method="class")
# Alternative way of building a classification tree
# cart1 <- rpart(as.factor(rev)~petit+respon+circuit+lctdir+issue+unconst,data=train)
```

## What's inside a CART?

The variable ```cart1``` contains all the information about the tree, with nodes, branches and the corresponding probability associated to each class. To analyze its content, we have a couple of options. 

The first is to use the function ```print()```: for each node, we know: node number, variable used for splitting, number of observations in the node, loss error in the node, predicted value, and predicted probabilities. Note that the children of node x are numbered 2x and 2x+1. Also, note that a variable might appear many times in the tree. The symbol * indicates a terminal node.

```{r, message=FALSE}
print(cart1)
```

The second option we have is to use the function ```summary()```.

```{r, results=FALSE}
summary(cart1)
```

## Visualizing a CART's structure

A fundamental advantage of CARTs is that it possible to visualize their structure. There is a number of packages that can be used to this purpose.

We begin with the default function (available in the ```rpart package```)

```{r}
plot(cart1); text(cart1) 
```

The `rpart` package has better functionalities

```{r, message=FALSE}
if(!require(rpart.plot)){   
  install.packages("rpart.plot")
  library(rpart.plot)
}
prp(cart1) 
prp(cart1,type=1) # labels all nodes (not just the leaves)
prp(cart1,type=4) # draws separate labels for left and right directions for all nodes and label nodes
prp(cart1,type=4,extra=4) # in addition, this also plots the probability per class of observation
# prp(cart1,type=4,extra=9) # probablities times fraction of observations at the node (the sum across all leaves is 1)
```

Other visualization options: see an example with the ```fancyRpartPlot()``` function, which requires ```rpart.plot```, ```rattle```, and ```RColorBrewer```.

```{r, message=FALSE}
if(!require(rattle)){
  install.packages("rattle")
  library(rattle)
}
if(!require(RColorBrewer)){
  install.packages("RColorBrewer")
  library(RColorBrewer)
}
fancyRpartPlot(cart1) # Run fancyRpartPlot
```

## Making predictions

The function ```predict()``` uses ```cart1``` and the test data to carry out the predictions. The function chooses the class with the highest probability.

```{r}
predictcart1 <- predict(cart1,newdata=test,type="class")
# predictcart1
```

How does the model perform?

```{r}
table(test$rev,predictcart1)
```

So the accuracy is ```r sum(diag(table(test$rev,predictcart1)))/sum(table(test$rev,predictcart1))```. 

Let's plot the ROC Curve:

```{r, message=FALSE}
predictcart1_prob <- predict(cart1,newdata=test) # predicted probabilities of rev=0 and rev=1 for each test observation
pred_cart1 <- prediction(predictcart1_prob[,2],test$rev)
perf_cart1 <- performance(pred_cart1, x.measure="fpr",measure="tpr")
plot(perf_cart1)
# performance(pred_cart1,measure="auc") 
as.numeric(performance(pred_cart1,measure="auc")@y.values)
```

Finally, we observe that the performance of the logistic regression and classification tree (CART) model are very similar.








