---
title: "Decision trees (Part 2)"
author: "Bikramjit Das"
date: "2025"
output: 
 html_document: 
    toc: true
    toc_depth: 2
    toc_float: true
    highlight: tango
geometry: margin=1in
fontsize: 12pt
linkcolor: blue    
---

First, we prepare the working environment.

```{r}
rm(list=ls()) # Clear the environment
# setwd("...")  # Setup the working directory
```

# Data Preparation

## Load the data on the US Supreme Court

```{r, message=FALSE}
supreme <- read.csv("supreme.csv") # We have 623 observations and 20 variables
# str(supreme)      # Internal structure of the dataframe 
# head(supreme)     # First part of the dataframe
# summary(supreme)  # Summary of the data
```

Recall the meaning of each variable:

-   `docket`: case number
-   `term`: year the case was discussed
-   `party_1` and `party_2`: parties involved in the case
-   `rehndir`, `stevdir`, `ocondir`, `scaldir`, `kendir`, `soutdir`, `thomdir`, `gindir`, `brydir`: direction of the judgement of the 9 judges (i.e., Rehnquist, Stevens, O'Connor, Scalia, Kennedy, Souter, Thomas, Ginsburg, Breyer). 0 means liberal, 1 conservative. 9 indicates that the vote is not available for the case.
-   `petit`: petitioner type
-   `respon`: respondant type
-   `circuit`: circuit of origin
-   `uncost`: binary number indicating if the petitioner argued practice was unconstitutional
-   `lctdir`: lower court direction (liberal or conservative)
-   `issue`: area of issue
-   `result`: 0 means liberal, 1 conservative

## Data for modelling justice Stevens' decisions

Let's now focus on a specific judge, say Stevens. We remove data with no entry from Stevens and process the variable `result`. This creates a new variable (`$rev`) in the dataframe that takes a value of 1 if Stevens' decision reverses the lower court decision and 0 if the judge affirms the lower court decision.

```{r}
stevens <- subset(supreme[,c("docket","term","stevdir","petit","respon","circuit","unconst","lctdir","issue","result")],supreme$stevdir!=9)
stevens$rev <- as.integer((stevens$lctdir=="conser" & stevens$stevdir==0) | (stevens$lctdir=="liberal" & stevens$stevdir==1))
```

Finally, we split the data into training and testing datasets. To this purpose, we use the `caTools` package.

```{r, message=FALSE}
if(!require(caTools)){                          # install or load the package
  install.packages("caTools")
  library(caTools)
}
set.seed(123)                                   # set seed for random sampling
spl <- sample.split(stevens$rev,SplitRatio=0.7) # We use 70% of the data for training
train <- subset(stevens,spl==TRUE);             # training dataset
test <- subset(stevens,spl==FALSE);             # testing dataset
```

# Warm up: Logistic regression

```{r}
m1 <- glm(rev~petit+respon+circuit+unconst+issue+lctdir,data=train,family="binomial")
# summary(m1)
```

Let's now try to make a prediction on the test data. We cannot use the command `p1 <- predict(m1,newdata=test,type="response")`, because `issue` has only one realization of the IR value (Interstate Relations), which is not estimated in the training set. Let's thus modify the test dataset as follows:

```{r}
test <- subset(test,test$issue!="IR")
p1 <- predict(m1,newdata=test,type="response")
```

How does the model perform?

```{r}
table(p1>=0.5,test$rev)
```

So the accuracy is `r sum(diag(table(p1>=0.5,test$rev)))/sum(table(p1>=0.5,test$rev))`.

# CART

For Classification And Regression Trees (CARTs) we first, we must load the `rpart` package.

```{r, message=FALSE}
if(!require(rpart)){        # rpart is for CART
  install.packages("rpart")
  library(rpart)
}
```

## Building a CART

The command `cart1 <- rpart(rev~petit+respon+circuit+lctdir+issue+unconst,data=train)` will not work, because `rpart` builds a regression tree by default. So, remember to specify you are working on a classification problem!

```{r}
cart1 <- rpart(rev~petit+respon+circuit+lctdir+issue+unconst,data=train,method="class")
# Alternative way of building a classification tree
# cart1 <- rpart(as.factor(rev)~petit+respon+circuit+lctdir+issue+unconst,data=train)
```

## What's inside a CART?

The variable `cart1` contains all the information about the tree, with nodes, branches and the corresponding probability associated to each class. To analyze its content, we have a couple of options.

The first is to use the function `print()`: for each node, we know: node number, variable used for splitting, number of observations in the node, loss error in the node, predicted value, and predicted probabilities. Note that the children of node x are numbered 2x and 2x+1. Also, note that a variable might appear many times in the tree. The symbol \* indicates a terminal node.

```{r, message=FALSE}
print(cart1)
```

The second option we have is to use the function `summary()`.

```{r, results=FALSE}
summary(cart1)
```

## Visualizing a CART's structure

A fundamental advantage of CARTs is that it possible to visualize their structure. There is a number of packages that can be used to this purpose.

We begin with the default function (available in the `rpart package`)

```{r}
plot(cart1); text(cart1) 
```

The `rpart` package has better functionalities

```{r, message=FALSE}
if(!require(rpart.plot)){   
  install.packages("rpart.plot")
  library(rpart.plot)
}
prp(cart1)        # default visualization
prp(cart1,type=1) # labels all nodes (not just the leaves)
prp(cart1,type=4) # draws separate labels for left and right directions for all nodes and label nodes
prp(cart1,type=4,extra=4) # in addition, this also plots the probability per class of observation
# prp(cart1,type=4,extra=9) # probabilities times fraction of observations at the node (the sum across all leaves is 1)
```

Other visualization options: see an example with the `fancyRpartPlot()` function, which requires `rpart.plot`, `rattle`, and `RColorBrewer`.

```{r, message=FALSE}
if(!require(rattle)){
  install.packages("rattle")
  library(rattle)
}
if(!require(RColorBrewer)){
  install.packages("RColorBrewer")
  library(RColorBrewer)
}
fancyRpartPlot(cart1) # Run fancyRpartPlot
```

## Making predictions

The function `predict()` uses `cart1` and the test data to carry out the predictions. The function chooses the class with the highest probability.

```{r}
predictcart1 <- predict(cart1,newdata=test,type="class")
# predictcart1
```

How does the model perform?

```{r}
table(test$rev,predictcart1)
```

So the accuracy is `r sum(diag(table(test$rev,predictcart1)))/sum(table(test$rev,predictcart1))`.

In a nutshell, we can see that the performance of the logistic regression and CART model are similar.

# Pruning

Let's begin by printing the `CP` table for model `cart1`. Note that these results are generated automatically by the `rpart` function using a 10-fold cross-validation.

```{r}
printcp(cart1)
```

Remarks:

-   `CP` is the Complexity Parameter ($\alpha$ in our slides)
-   We pick the value of `CP` that minimizes cross-validation error `xerror` $\pm$ its standard deviation `xstd`
-   In our case, it is clear that we achieve the best performance for `CP` equal to 0.030769
-   The errors are scaled so that the first node has error = 1.
-   The `CP` parameter is also scaled: the default stopping value used in `rpart` is 0.01.
-   (Roughly, the relative error can be converted to absolute error by multiplying the Root node error by the `rel error`)

We can also plot the relation between `CP` and `error`:

```{r}
plotcp(cart1)
```

Let's now prune the tree

```{r}
opt <- which.min(cart1$cptable[,"xerror"]) # get index of CP with lowest xerror
cp <- cart1$cptable[opt, "CP"]            # get the corresponding value
cart2 <- prune(cart1,cp)
```

As shown below, this is a much smaller tree!

```{r}
prp(cart2,type=4,extra=4)
```

And yet, its accuracy on the test dataset is higher than the one of a fully-grown tree!

```{r}
predictcart2 <- predict(cart2,newdata=test,type="class")
table(test$rev,predictcart2)
```

The accuracy is `r sum(diag(table(test$rev,predictcart2)))/sum(table(test$rev,predictcart2))`.
