---
title: "Multiple testing"
author: "Bikramjit Das"
date: "2025"
output: 
 html_document: 
    toc: true
    toc_depth: 2
    toc_float: true
    highlight: tango
geometry: margin=1in
fontsize: 12pt
linkcolor: blue    
---

First, we prepare the working environment.
```{r}
rm(list=ls())   # Clear the environment
# setwd("...")  # Setup the working directory
```






## Review of Hypothesis Tests

### One-sample $t$-test
We begin by performing some one-sample $t$-tests using the `t.test()` function. 

* First we create 100 variables, each consisting of 10 observations. 

* The first 50 variables have mean $\mu_j=0.5, i=j,\ldots, 50$ and variance 1, while the others have mean $\mu_j=0, j=51, \ldots, 100$ and variance  1.


```{r}
set.seed(6)
x <- matrix(rnorm(10 * 100), 10, 100)
x[, 1:50] <- x[, 1:50] + 0.5
```

* The `t.test()` function can perform a one-sample or a two-sample $t$-test.
By default, a one-sample test is performed.

* To begin, we test $\textbf{H}_0 : \mu_1 = 0.$


```{r}
t.test(x[, 1], mu = 0)
```


* The *p-value* comes out to 0.067, which is not quite low enough to reject
the null hypothesis at level $\alpha = 0.05$. 

* In this case, $\mu_1=0.5$, so the null hypothesis is false. Therefore, we have made a Type II error by failing to
reject the null hypothesis when the null hypothesis is false.

### Multiple $t$-tests

* We now test $\textbf{H}_{0j} : \mu_j = 0.$ for $j = 1, . . . , 100$. 

* We compute the 100 *p-values*, and then construct a vector recording whether the $j$-th *p-value* is less than or equal to 0.05, in which case we reject $\textbf{H}_{0j}$ , or greater than 0.05, in which case we do not reject $\textbf{H}_{0j}$ , for $j = 1, . . . , 100$. 

* Then we create a table to check the correctness of our tests.

```{r}
p.values <- rep(0, 100)
for (i in 1:100)
  p.values[i] <- t.test(x[, i], mu = 0)$p.value
decision <- rep("Do not reject H0", 100)
decision[p.values <= .05] <- "Reject H0"
## Table
table(decision, c(rep("H0 is False", 50), rep("H0 is True", 50)))
```
* Therefore, at level $\alpha=0.05$, we reject just 10 of the 50 false null hypotheses, and we incorrectly reject 3 of the true null hypotheses. 

* Using the notation from class, we have $W = 40, U = 47, S = 10, V = 3$. 
We have set $\alpha=0.05$, which means that we expect to reject around 5% of
the true null hypotheses. This is in line with the 2 × 2 table above, which
indicates that we rejected $V = 3$ of the 50 true null hypotheses.

* In the simulation above, for the false null hypotheses, the ratio of the
mean to the standard deviation was only 0.5/1 = 0.5. This amounts to
quite a weak signal, and it resulted in a high number of Type II errors. If
we instead simulate data with a stronger signal, so that the ratio of the
mean to the standard deviation for the false null hypotheses equals 1, then
we make only 9 Type II errors as seen below.
```{r}
x <- matrix(rnorm(10 * 100), 10, 100)
x[, 1:50] <- x[, 1:50] + 1
for (i in 1:100)
  p.values[i] <- t.test(x[, i], mu = 0)$p.value
decision <- rep("Do not reject H0", 100)
decision[p.values <= .05] <- "Reject H0"
table(decision,
      c(rep("H0 is False", 50), rep("H0 is True", 50))
)
```



## The Family-Wise Error Rate

Now we test for $m$ independent hypotheses and check the family-wise  rate  (FWER)
for different levels $\alpha=0.001, 0.01, 0.05$ as $m$ increases from 1 to 500.
```{r}
m <- 1:500
fwe1 <- 1 - (1 - 0.05)^m
fwe2 <- 1 - (1 - 0.01)^m
fwe3 <- 1 - (1 - 0.001)^m
### Plotting
par(mfrow = c(1, 1))
plot(m, fwe1, type = "l", log = "x", ylim = c(0, 1), col = 2,
     ylab = "Family - Wise Error Rate",
     xlab = "Number of Hypotheses")
lines(m, fwe2, col = 4)
lines(m, fwe3, col = 3)
abline(h = 0.05, lty = 2)

```

* Even for moderate values of $m$ such as 50, the FWER exceeds 0.05 unless $\alpha$ is set to a very low value, such as 0.001. 

* The problem with setting $\alpha$ to such a low value is that we are likely to make a number of Type II errors.

### Fund manager dataset

We now conduct a one-sample $t$-test for each of the first five managers
in the `fund.mini` dataset, in order to test the null hypothesis that the $j$-th fund manager’s mean return equals zero, $\textbf{H}_{0j} : \mu_j = 0.$
```{r}
library(ISLR2)
fund.mini <- Fund[, 1:5]
t.test(fund.mini[, 1], mu = 0)
fund.pvalue <- rep(0, 5)
for (i in 1:5)
  fund.pvalue[i] <- t.test(fund.mini[, i], mu = 0)$p.value
fund.pvalue
```

* The *p-values* are low for Managers One and Three, and high for the other
three managers. However, we cannot simply reject $\textbf{H}_{01}$ and $\textbf{H}_{03}$, since this would fail to account for the multiple testing that we have performed.
 
* Instead, we will conduct Bonferroni’s method and Holm’s method to control
the FWER.


* To do this, we use the `p.adjust()` function. Given the *p-values*, the function outputs *adjusted  p-values*, which can be thought of as a new set of  *p-values* that have been corrected for multiple testing. If the adjusted *p-value*  for a given hypothesis is less than or equal to $\alpha$, then that hypothesis can be rejected while maintaining a FWER of no more than $\alpha$. In other
words, the adjusted *p-values* resulting from the `p.adjust()` function can
simply be compared to the desired FWER in order to determine whether
or not to reject each hypothesis.

* For example, in the case of Bonferroni’s method, the raw *p-values* are
multiplied by the total number of hypotheses, $m$, in order to obtain the
adjusted *p-values*. (However, adjusted *p-values* are not allowed to exceed 1.)


```{r}
p.adjust(fund.pvalue, method = "bonferroni")
pmin(fund.pvalue * 5, 1)
```

* Therefore, using Bonferroni’s method, we are able to reject the null hypothesis only for Manager One while controlling the FWER at 0.05.

* By contrast, using Holm’s method, the adjusted p-values indicate that
we can reject the null hypotheses for Managers One and Three at a FWER
of 0.05.

```{r}
p.adjust(fund.pvalue, method = "holm")
```

#### Pair-wise comparison

Notice that Manager One seems to perform particularly well,
whereas Manager Two has poor performance. 
```{r}
apply(fund.mini, 2, mean)
#apply(fund.mini, 2, hist)
```

Is there evidence of a meaningful difference in performance between these two managers? 

* Performing a paired  $t$-test  using the `t.test()` function results in a *p-value*  of 0.038, suggesting a statistically significant difference.
```{r}
t.test(fund.mini[, 1], fund.mini[, 2], paired = T)
```


Pairwise comparison test using Tukey's method can be conducted which we skip here.



## The False Discovery Rate

Now we perform hypothesis tests for all 2,000 fund managers in the `fund`
dataset. We perform a one-sample  $t$-test  of $\textbf{H}_{0j}: \mu_j = 0$, which states that the $j$-th fund manager’s mean return is zero.

```{r}
fund.pvalues <- rep(0, 2000)
for (i in 1:2000)
  fund.pvalues[i] <- t.test(Fund[, i], mu = 0)$p.value
```

* There are far too many managers to consider trying to control the FWER.
Instead, we focus on controlling the FDR: that is, the expected fraction of
rejected null hypotheses that are actually false positives. 

* The `p.adjust()` function can be used to carry out the Benjamini-Hochberg procedure.

```{r}
q.values.BH <- p.adjust(fund.pvalues, method = "BH")
q.values.BH[1:10]
```

* The *q-values* output by the Benjamini-Hochberg procedure can be interpreted as the smallest FDR threshold at which we would reject a particular null hypothesis. 

* For instance, a *q-value* of 0.1 indicates that we can reject
the corresponding null hypothesis at an FDR of 10% or greater, but that
we cannot reject the null hypothesis at an FDR below 10%.

* If we control the FDR at 10%, then for how many of the fund managers
can we reject $\textbf{H}_{0j}: \mu_j = 0$?

```{r}
sum(q.values.BH <= 0.3)
```

* We find that 146 of the 2,000 fund managers have a *q-value* below 0.1;
therefore, we are able to conclude that 146 of the fund managers beat the
market at an FDR of 10%. Only about 15 (10% of 146) of these fund
managers are likely to be false discoveries. 

* By contrast, if we had instead
used Bonferroni’s method to control the FWER at level $\alpha = 0.1$, then we would have failed to reject any null hypotheses!
```{r}
sum(fund.pvalues <= (0.1 / 2000))
```


* Recall that the Benjamini-Hochberg procedure searches
for the largest *p-value*  such that $p(j) < qj/m$, and rejects all hypotheses
for which the *p-value*  is less than or equal to $p(j)$. 

* In the code below, we implement the Benjamini-Hochberg procedure ourselves, in order to illustrate how it works. 

```{r}
ps <- sort(fund.pvalues)
m <- length(fund.pvalues)
q <- 0.1
wh.ps <- which(ps < q * (1:m) / m)
if (length(wh.ps) >0) {
  wh <- 1:max(wh.ps)
} else {
  wh <- numeric(0)
}
```

* We plot the ordered *p-values* against their indices for $\alpha=0.1.$ The green line is the FWER control and the orange line the FDR control.
```{r}
plot(ps, log = "xy", ylim = c(4e-6, 1), ylab = "p-value",
     xlab = "Index", main = "")
points(wh, ps[wh], col = 4)
abline(a = 0, b = (q / m), col = 2, untf = TRUE)
abline(h = 0.1 / 2000, col = 3)
```


## A Re-Sampling Approach

Here, we implement the re-sampling approach to hypothesis testing using
the `Khan` dataset. 

* First, we merge the training and testing data, which results in observations on 83 patients for 2,308 genes.

```{r}
attach(Khan)
x <- rbind(xtrain, xtest)
y <- c(as.numeric(ytrain), as.numeric(ytest))
dim(x)
table(y)
```

* There are four classes of cancer. For each gene, we compare the mean expression in the second class (rhabdomyosarcoma) to the mean expression in
the fourth class (Burkitt’s lymphoma). 

* Performing a standard two-sample $t$-test on the 11th gene produces a test-statistic of −2.09 and an associated *p-value*  of 0.0412, suggesting modest evidence of a difference in mean expression levels between the two cancer types.
```{r}
x <- as.matrix(x)
x1 <- x[which(y == 2), ]
x2 <- x[which(y == 4), ]
n1 <- nrow(x1)
n2 <- nrow(x2)
t.out <- t.test(x1[, 11], x2[, 11], var.equal = TRUE)
TT <- t.out$statistic
TT
t.out$p.value
```

* However, this *p-value*  relies on the assumption that under the null hypothesis of no difference between the two groups, the test statistic follows a $t$-distribution with 29 + 25 − 2 = 52 degrees of freedom. 

* Instead of using this theoretical null distribution, we can randomly split the 54 patients into two groups of 29 and 25, and compute a new test statistic. Under the null hypothesis of no difference between the groups, this new test statistic should have the same distribution as our original one. 

* Repeating this process 10,000 times allows us to approximate the null distribution of the test statistic. We compute the fraction of the time that our observed test statistic exceeds the test statistics obtained via re-sampling.
```{r}
set.seed(1)
B <- 10000
Tbs <- rep(NA, B)
for (b in 1:B) {
  dat <- sample(c(x1[, 11], x2[, 11]))
  Tbs[b] <- t.test(dat[1:n1], dat[(n1 + 1):(n1 + n2)],
                   var.equal = TRUE
  )$statistic
}
mean((abs(Tbs) >= abs(TT)))
```
* This fraction, 0.0416, is our re-sampling-based $p$-value. It is almost identical
to the *p-value*  of 0.0412 obtained using the theoretical null distribution.

* We can plot a histogram of the re-sampling-based test statistics.

```{r}
hist(Tbs, breaks = 100, xlim = c(-4.2, 4.2), main = "",
     xlab = "Null Distribution of Test Statistic", col = 7)
lines(seq(-4.2, 4.2, len = 1000),
      dt(seq(-4.2, 4.2, len = 1000),
         df = (n1 + n2 - 2)
      ) * 1000, col = 2, lwd = 3)
abline(v = TT, col = 4, lwd = 2)
text(TT + 0.5, 350, paste("T = ", round(TT, 4), sep = ""),
     col = 4)
```

* The re-sampling-based null distribution is almost identical to the theoretical
null distribution, which is displayed in red.

* Finally, we implement the plug-in re-sampling FDR approach.

* Calculating
the FDR for all 2,308 genes in the Khan dataset may take a while. Hence,
we will illustrate the approach on a random subset of 100 genes. For each
gene, we first compute the observed test statistic, and then produce 10,000
re-sampled test statistics. This may take a few minutes to run.


```{r}
m <- 100
set.seed(1)
index <- sample(ncol(x1), m)
Ts <- rep(NA, m)
Ts.star <- matrix(NA, ncol = m, nrow = B)
for (j in 1:m) {
  k <- index[j]
  Ts[j] <- t.test(x1[, k], x2[, k],
                  var.equal = TRUE
  )$statistic
  for (b in 1:B) {
    dat <- sample(c(x1[, k], x2[, k]))
    Ts.star[b, j] <- t.test(dat[1:n1],
                            dat[(n1 + 1):(n1 + n2)], var.equal = TRUE
    )$statistic
  }
}
```
* Next, we compute the number of rejected null hypotheses R, the estimated
number of false positives $\widehat{V}$, and the estimated FDR, for a range of threshold
values $c$. The threshold values are chosen using the
absolute values of the test statistics from the 100 genes.
```{r}
cs <- sort(abs(Ts))
FDRs <- Rs <- Vs <- rep(NA, m)
for (j in 1:m) {
  R <- sum(abs(Ts) >= cs[j])
  V <- sum(abs(Ts.star) >= cs[j]) / B
  Rs[j] <- R
  Vs[j] <- V
  FDRs[j] <- V / R
}
```
* Now, for any given FDR, we can find the genes that will be rejected. For
example, with the FDR controlled at 0.1, we reject 15 of the 100 null
hypotheses. 

* On average, we would expect about one or two of these genes
(i.e. 10% of 15) to be false discoveries. 

* At an FDR of 0.2, we can reject the null hypothesis for 28 genes, of which we expect around six to be
false discoveries. 

* The variable index is needed here since we restricted our
analysis to just 100 randomly selected genes.
```{r}
max(Rs[FDRs <= .1])
sort(index[abs(Ts) >= min(cs[FDRs < .1])])
max(Rs[FDRs <= .2])
sort(index[abs(Ts) >= min(cs[FDRs < .2])])
```

*This material is adapted from the Lab tutorial in **An Introduction to Statistical Learning with Applications in R, Second edition, Chapter 13**.*
