---
title: "Social Progress Index with PCA"
author: "Bikramjit Das"
date: "2025"
output: 
 html_document: 
    toc: true
    toc_depth: 2
    toc_float: true
    highlight: tango
geometry: margin=1in
fontsize: 12pt
linkcolor: blue  

---

First, we prepare the working environment.
```{r}
rm(list=ls()) # Clear the environment
# setwd("...")  # Setup the working directory
# include ggplot if required
if(!require(ggplot2)){
  install.packages("ggplot2")
  library(ggplot2)
}
## install factoextra
#if(!require(factoextra)){
 # install.packages("factoextra")
#  library(factoextra)
#}

```


# Introduction


**New York Times opinion column:**

https://www.nytimes.com/2020/09/09/opinion/united-states-social-progress.html 

**Social Progress Imperative** 

https://www.socialprogress.org/

Quoting from their report: The Social Progress Index is a well-established measure, published since 2013, that is meant to catalyze improvement and drive action by presenting social outcome data in a useful and reliable way. Composed of multiple dimensions, the Social Progress Index can be used to benchmark success and provide a holistic, transparent, outcome-based measure of a country’s well-being that is independent of economic indicators. Policymakers, businesses, and countries’ citizens alike can use it to compare their country against others on different facets of social progress, allowing the identification of specific areas of strength or weakness. 

We look at a recent social progress data from 2011-2020. Among the 192 countries listed in 2020, 163 are ranked according to Social Progress Index. We will consider this data set for our study.


## Motivation
Can we come up with a reasonable index of social progress from the data that we have?

## Read data
```{r, results='hide'}

### Read data
sp2011to2020 <- read.csv("socprog2020.csv")
sp2011to2020
### data check
head(sp2011to2020)
dim(sp2011to2020)
```


Note that there is a ranking of the countries and a social progress index mentioned here.

## Data processing

We select the relevant parts for use. First we take the country information and remove first few columns. Keep year 2020 and *Ranked* ones.

```{r, results='hide'}
spi2020<-subset(sp2011to2020,SPI.year =="2020" & Status=="Ranked") ## we only consider the year 2020 and countries which were ranked
rownames(spi2020)<-spi2020[,2] ## he row names are now country names

spirank<-spi2020[,1]    #### ranks of the countries. to be used later
spi2020<-spi2020[,-c(1:6)]
dim(spi2020)
#spi2020
```

Then we remove columns which do not have available data (remove "NA").


```{r,results='hide'}
NAobject<-colnames(spi2020)[apply(is.na(spi2020),2,any)]
spi2020<-spi2020[,!names(spi2020)%in% NAobject]
### check
dim(spi2020)
head(spi2020)
```

## Singapore
How is Singapore fairing here? What about the US?

```{r, results='hide'}
#rownames(spi2020)
spi2020["Singapore",]
spi2020["United States",]
summary(spi2020$Access.to.Basic.Knowledge)
spi2020[which.max(spi2020$Access.to.Basic.Knowledge),]
```

We can also compute average over all countries for the different attributes.
```{r,results='hide'}
#names(spi2020)
 Stats<-as.data.frame(cbind(apply(spi2020, 2, mean),apply(spi2020, 2, "median"),apply(spi2020, 2, "sd")))
 colnames(Stats)<-c("mean", "median","std.dev")

 head(Stats)
# View(Stats) ## note that they are on different scales
```

# PCA without scaling
Let us do some basic principal component analysis for social progress. 
```{r, results='hide'}
?prcomp
pr.out.ns=prcomp(spi2020, scale=F)
pr.out.ns$rotation[,1]
a<-pr.out.ns$rotation[,1:3]     ### gives the weights for the PCAs (eigen vectors)
pr.out.ns$x[1:10,1:5]  ### gives PCA value for 1st 10 countries and first 5 PCAs
```

## Scree plot
```{r}
#plot scree: plot of variance of each PCA
plot(pr.out.ns, type="l", main="Scree plot")

#plot cumulative prop: proportional variance (cumulative)

pr.var.ns=pr.out.ns$sdev^2
pve.ns=pr.out.ns$sdev^2/sum(pr.out.ns$sdev^2)
plot(cumsum(pve.ns), xlab="Principal Component", ylab="Cumulative Proportion of Variance Explained", ylim=c(0,1),type='b')+abline(h=0.9,col="red")
# first PC explanation
pve.ns[1]
```

From the `Scree plot' and the 'Cumulative Proportion plot': The first PC actually explains 93% of the variability.

Let us look at the loadings for each and the PCA scores for the countries. We look at the coefficients of the top and bottom 5. Also we also look at the top and bottom scoring countries for each PC.

## Significant loadings, country rankings 

```{r, results='hide'}
#pr.out.ns$x
#pr.out.ns$rotation
sp.pc<-data.frame(pr.out.ns$x[,1:2])
sp.pc[order(-sp.pc$PC1),]
sp.pc[order(-sp.pc$PC2),]
dim(spi2020)


co.sp.pc<-data.frame(pr.out.ns$rotation[,1:2])

View(co.sp.pc)

View(co.sp.pc[order(-co.sp.pc[,1]),][c(1:5,50:54),])
View(co.sp.pc[order(-co.sp.pc[,2]),][c(1:5,50:54),])

View(Stats)
``` 

We observe that PC1 explains 93% of the variation and moreover PC1 is dominated by Green house gas emission. So countries like China, US, India, Brazil are clustered and dominate together. On the other hand PC2 seems dominated by mortality rates from infectious disease, child birth and other premature deaths. 

# PCA with scaling

Clearly, these are the components with very high variance anyway. Hence what is often suggested here is to standardize or scale the data.

## Scree plot
```{r, results='hide'}
pr.out=prcomp(spi2020, scale=TRUE)
plot(pr.out, type="l", main="Scree plot")
plot(pr.out, type="l", main="Scree plot")+abline(h=1,col="red")
pr.var=pr.out$sdev^2
pr.var
pve=pr.var/sum(pr.var)
pve
plot(pve, xlab="Principal Component", ylab="Proportion of Variance Explained", ylim=c(0,1),type='b')
plot(cumsum(pve), xlab="Principal Component", ylab="Cumulative Proportion of Variance Explained", ylim=c(0,1),type='b')  #+ abline(h=0.8,col="red")+abline(v=6,col="blue")
```

## Significant loadings, country rankings 
```{r, results='hide'}
## Which components and which country?
 sp4<-data.frame(pr.out$x[,1:4])
 sp4[order(-sp4$PC1),]
 
 cosp4<-data.frame(pr.out$rotation[,1:4])
 cosp4[order(-cosp4[,1]),]
``` 








## Correlation plot 
```{r}
 ## Correlation plot
 ## Check Stats earlier
 library(corrplot)
 corspi<-cor(spi2020)
 dim(corspi)
 rownames(corspi)<-colnames(corspi)<-1:54
 corrplot(corspi,type="upper")
```


# PCA with only the core quantities
Now, we can go back to the documentation of the data and see that the primary components here are not the core metrics computed. There are three big dimensions: 
1. Basic Human Needs
2. Foundations of Well-being
3. Opportunity

The SPI is an average of these three quantities which are again averages of four different quantities each. We remove this to have the core data to study.
 
```{r}
  bigdim<-c("Basic.Human.Needs","Foundations.of.Wellbeing","Opportunity")
 medcomp<-c("Nutrition.and.Basic.Medical.Care","Water.and.Sanitation","Shelter","Personal.Safety","Access.to.Basic.Knowledge","Access.to.Information.and.Communications","Health.and.Wellness","Environmental.Quality","Personal.Rights","Personal.Freedom.and.Choice","Inclusiveness","Access.to.Advanced.Education")
 
 spi2020core<-spi2020[,!names(spi2020)%in% c(bigdim,medcomp)]
 dim(spi2020core)

 
 corspicore<-cor(spi2020core)
 rownames(corspicore)<-colnames(corspicore)<-1:dim(corspicore)
```

## Correlation plot
```{r} 
 corrplot(corspicore,type="upper")
```

 Now we conduct a PCA.
  What we observe is that the PC1 explains 50.0%
                              PC2 explains 14.2%
                              PC3 explains  4.6%  
                              PC4 explains  3.7%
```{r}
pr.out.core=prcomp(spi2020core, scale=TRUE)
#pr.out.core
plot(pr.out.core, type="l", main="Scree plot for core indicators") +abline(h=1,col="red")

pr.var.core=pr.out.core$sdev^2
#pr.var.core
pve.core=pr.var.core/sum(pr.var.core)
pve.core
plot(pve.core, xlab="Core indicators: Principal Component", ylab="Proportion of Variance Explained", ylim=c(0,1),type='b')
plot(cumsum(pve.core), xlab="Core indicators: Principal Component", ylab="Cumulative Proportion of Variance Explained", ylim=c(0,1),type='b')
```

```{r, results='hide'}
n<-dim(spi2020core)[1]
k<-dim(spi2020core)[2]

nup<-5
kup<-5


 sp4.core<-data.frame(pr.out.core$x[,1:4])
 cosp4.core<-data.frame(pr.out.core$rotation[,1:4])
 #sp4.core
 #cosp4.core
 
  

 cosp4.core[order(-cosp4.core[,1]),][c(1:kup,(k-kup+1):k),]
 topbotPC1<-row.names(cosp4.core[order(-cosp4.core[,1]),])[c(1:kup,(k-kup+1):k)]
 sp4.core[order(-sp4.core$PC1),][c(1:nup,(n-nup+1):n),]
 topbotind1<-row.names(sp4.core[order(-sp4.core$PC1),])[c(1:nup,(n-nup+1):n)]
 
 
 cosp4.core[order(-cosp4.core[,2]),][c(1:kup,(k-kup+1):k),]
 topbotPC2<-row.names(cosp4.core[order(-cosp4.core[,2]),])[c(1:kup,(k-kup+1):k)]
 sp4.core[order(-sp4.core$PC2),][c(1:nup,(n-nup+1):n),]
 topbotind2<-row.names(sp4.core[order(-sp4.core$PC2),])[c(1:nup,(n-nup+1):n)]
 
 
 cosp4.core[order(-cosp4.core[,3]),][c(1:kup,(k-kup+1):k),]
 topbotPC3<-row.names(cosp4.core[order(-cosp4.core[,3]),])[c(1:kup,(k-kup+1):k)]
 sp4.core[order(sp4.core$PC3),][c(1:nup,(n-nup+1):n),]
 topbotind3<-row.names(sp4.core[order(-sp4.core$PC3),])[c(1:nup,(n-nup+1):n)]
 
 
 cosp4.core[order(-cosp4.core[,4]),][c(1:kup,(k-kup+1):k),]
 topbotPC4<-row.names(cosp4.core[order(-cosp4.core[,3]),])[c(1:kup,(k-kup+1):k)]
 sp4.core[order(-sp4.core$PC4),][c(1:nup,(n-nup+1):n),]
 topbotind4<-row.names(sp4.core[order(-sp4.core$PC4),])[c(1:nup,(n-nup+1):n)]

 # Check that there is no correlation.
 #corspipc<-cor(pr.out.core$x)
 #corrplot(corspipc,type="upper")
 
```
 
 
1. PC1: Foundations of well-being, health and wellness (+), ... Basic human needs: lack of water and sanitation/shelter  (-) 

2. PC2: Opportunity, lack of personal rights (-), ... Foundations of well-being, environment quality (+)

3. PC3: ...


# Plotting the component loadings and the individuals

## This part is optional and will not be tested

Now we do some plotting of component loadings and the individual countries along PCAs. We use soemthing called a biplot. The library `factoextra` is used here.
 
```{r}
library(factoextra)
 #fviz_eig(pr.out)
 fviz_pca_ind(pr.out.core, labelsize=3,
             col.ind = "contrib", # Color by the quality of representation
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             geom = c("point","text"),
             #select.ind = list(cos2 = 30),
             select.ind= list(name=c(topbotind1,topbotind2)),
             repel = TRUE     # Avoid text overlapping
             )
 
 ggsave("CountryPC1PC2.png")
 
 fviz_pca_var(pr.out.core, labelsize=3,
             col.var = "contrib", # Color by contributions to the PC
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             select.var = list(name=c(topbotPC1,topbotPC2)),
             repel = TRUE     # Avoid text overlapping
             )
 
  ggsave("PC1PC2.png")
  

  fviz_pca_ind(pr.out.core, labelsize=3,
             col.ind = "contrib", # Color by the quality of representation
             axes=c(3,4),
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             geom = c("point","text"),
             #select.ind = list(cos2 = 30),
             #theme(text = element_text(size = 2)),
             select.ind= list(name=c(topbotind3,topbotind4)),
             repel = TRUE     # Avoid text overlapping
             )
  
   ggsave("CountryPC3PC4.png")

 fviz_pca_var(pr.out.core, labelsize=3,
             col.var = "cos2", # Color by contributions to the PC
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             axes=c(3,4),
             select.var = list(name=c(topbotPC3,topbotPC4)),
             repel = TRUE     # Avoid text overlapping
             )
  ggsave("PC3PC4.png")

```

# An index of our own
 
Can we come up with a rank based on the PCAs. We can just use the first PCA. In general it is not suggested that we add scores from different PCAs ... but if we still go ahead and do that for one more.

```{r}
spi_slds<-data.frame(sp4.core[,1])#-sp4.core[,2])#-sp4.core[,3]+sp4.core[,4])
rownames(spi_slds)<-rownames(sp4.core)
spi_slds$rank<-rank(-spi_slds)
spi_slds$spirank<-spirank
colnames(spi_slds)<-c("Score","Rank SLDS","Rank SPI")
spi_slds[order(spi_slds[,2]),][c(1:10,154:163),]
```

###### End of PCA Analysis for Social Progress data.
